---
title: "CSP_571_Project"
output: html_document
date: "2024-04-1"
---

## CSP 571 Project

### Data Preparation

#### Importing data

```{r}
library(readr)
path_file = "Crimes_-_2001_to_Present_20231109(1).csv"
crime_data <- read_csv(path_file)
head(crime_data)
```
#### Taking a subset of the file for ease of trials
```{r}
crime_data_2020_to_2023 <- subset(crime_data, Year >= 2020 & Year <= 2023)
write.csv(crime_data_2020_to_2023, file = "crime_data_2020_to_2023.csv", row.names = FALSE)
```
```{r}
library(readr)
crime_data_subset <- read_csv("crime_data_2020_to_2023.csv")
str(crime_data_subset)
```
#### Removing irrelevant features
```{r}
library(dplyr)
crime_data_subset <- select(crime_data_subset, Date, `Primary Type`, `Location Description`, District, `Community Area`, Arrest)
str(crime_data_subset)
```
#### Removing duplicates and null values
```{r}
library(dplyr)
crime_data_subset <- distinct(crime_data_subset)
crime_data_subset <- na.omit(crime_data_subset)
str(crime_data_subset)
```
#### Transforming date into seperate columns
```{r}
library(lubridate)
crime_data_subset$Date <- mdy_hms(crime_data_subset$Date)

crime_data_subset$Day <- day(crime_data_subset$Date)
crime_data_subset$Month <- month(crime_data_subset$Date)
crime_data_subset$Year <- year(crime_data_subset$Date)

crime_data_subset$Time <- case_when(
  hour(crime_data_subset$Date) >= 6 & hour(crime_data_subset$Date) < 12 ~ "Morning",
  hour(crime_data_subset$Date) >= 12 & hour(crime_data_subset$Date) < 18 ~ "Afternoon",
  hour(crime_data_subset$Date) >= 18 & hour(crime_data_subset$Date) < 24 ~ "Evening",
  TRUE ~ "Night"
)
crime_data_subset <- select(crime_data_subset, -Date)
head(crime_data_subset)
```
### Data Analysis

#### Getting the top 5 crimes
```{r}
# install.packages("ggplot2")
library(ggplot2)

crime_counts <- crime_data_subset %>%
  count(`Primary Type`, sort = TRUE)

total_crimes <- sum(crime_counts$n)
crime_counts$percentage <- (crime_counts$n / total_crimes) * 100

top_5_crimes <- head(crime_counts, 5)

pie(crime_counts$percentage, labels = paste(crime_counts$`Primary Type`, "\n", round(crime_counts$percentage, 1), "%"), col = rainbow(length(top_5_crimes$`Primary Type`)), main = "Occurrences of Crime", cex = 0.5)

ggplot(top_5_crimes, aes(x = reorder(`Primary Type`, -n), y = n, fill = `Primary Type`)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round((n / total_crimes) * 100, 1), "%")), vjust = -0.5, size = 3) +
  labs(title = "Top 5 Crimes", x = "Crime Type", y = "Occurrences") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = rainbow(length(top_5_crimes$`Primary Type`))) +
  guides(fill = FALSE)
```
#### Getting the data for the top 5 crimes
```{r}
top_5_crimes_data <- crime_data_subset %>%
  filter(`Primary Type` %in% top_5_crimes$`Primary Type`)
str(top_5_crimes_data)
```
#### Analysing the crimes
```{r}
arrest_rate <- top_5_crimes_data %>%
  group_by(`Primary Type`) %>%
  summarise(arrest_rate = mean(Arrest, na.rm = TRUE))

top_5_crimes_with_arrest_rate <- left_join(top_5_crimes, arrest_rate, by = "Primary Type")

arrest_rate_percentage <- top_5_crimes_with_arrest_rate$arrest_rate * 100
pie(arrest_rate_percentage, 
    labels = paste(top_5_crimes_with_arrest_rate$`Primary Type`, "\n", round(arrest_rate_percentage, 2), "%"),
    main = "Arrest Rates for Top 5 Crimes",
    col = rainbow(length(arrest_rate_percentage))
)
```

### Analysing spatial and temporal patterns for the top 5 crime types


### Analysing spatial and temporal patterns for the top 5 crime types


### Spatial Analysis

##### 1. Distribution of crimes by District and Community Area

```{r}
library(dplyr)
library(ggplot2)

# Define specific crimes to analyze
specific_crimes <- c("BATTERY", "THEFT", "MOTOR VEHICLE THEFT", "CRIMINAL DAMAGE", "ASSAULT")

# Analyze each crime type
for (crime in specific_crimes) {
  cat("\n--- Spatial Analysis for:", crime, "---\n")
  
  crime_data_specific <- crime_data_subset %>% filter(`Primary Type` == crime)
  
  # Distribution by District
  district_counts <- crime_data_specific %>%
    count(District) %>%
    mutate(District = factor(District))  # Ensuring District is treated as a factor for plotting

  plot_districts <- ggplot(district_counts, aes(x = District, y = n, fill = District)) +
    geom_bar(stat = "identity") +
    labs(title = paste("Crime Distribution by District for", crime), x = "District", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

  print(plot_districts)
}

```
##### 2. Arrest Rates by district and Community Area

```{r}
for (crime in specific_crimes) {
  cat("\n--- Spatial Analysis of Arrest Rates for:", crime, "---\n")
  
  crime_data_specific <- crime_data_subset %>% filter(`Primary Type` == crime)
  
  # Arrest Rates by District
  arrest_rate_district <- crime_data_specific %>%
    group_by(District) %>%
    summarise(Arrests = sum(Arrest, na.rm = TRUE),
              Total = n(),
              Arrest_Rate = (Arrests / Total) * 100) %>%
    ungroup()

  plot_arrest_district <- ggplot(arrest_rate_district, aes(x = reorder(District, -Arrest_Rate), y = Arrest_Rate, fill = District)) +
    geom_col() +
    labs(title = paste("Arrest Rate by District for", crime), x = "District", y = "Arrest Rate (%)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

  print(plot_arrest_district)
}
```
### Temporal Analysis

##### 1. Crime Trend Over Years

Code for Yearly Crime Trends:

```{r}
library(dplyr)
library(ggplot2)

# Analyze crime trends over years
yearly_trends <- crime_data_subset %>%
  count(Year, `Primary Type`, name = "Count") %>%
  filter(`Primary Type` %in% specific_crimes)

ggplot(yearly_trends, aes(x = Year, y = Count, color = `Primary Type`, group = `Primary Type`)) +
  geom_line() +
  geom_point() +
  labs(title = "Yearly Crime Trends by Crime Type", x = "Year", y = "Number of Crimes") +
  theme_minimal()
```


###### 2. Monthly Crime Distribution

Code for Monthly Crime Distribution:

```{r}
# Monthly distribution of crimes
monthly_distribution <- crime_data_subset %>%
  count(Month, `Primary Type`, name = "Count") %>%
  filter(`Primary Type` %in% specific_crimes)

ggplot(monthly_distribution, aes(x = Month, y = Count, fill = `Primary Type`)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Monthly Distribution of Crimes by Type", x = "Month", y = "Number of Crimes") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_minimal()
```

##### 3. Crime Frequency by Time of Day

Code for Crime Frequency by Time of Day:

```{r}
# Monthly distribution of crimes
monthly_distribution <- crime_data_subset %>%
  count(Month, `Primary Type`, name = "Count") %>%
  filter(`Primary Type` %in% specific_crimes)

ggplot(monthly_distribution, aes(x = Month, y = Count, fill = `Primary Type`)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Monthly Distribution of Crimes by Type", x = "Month", y = "Number of Crimes") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_minimal()
```






```{r}

```

#### Repeat for the rest 4 crimes

### Model Selection 
```{r}

```
#### Feature selection techniques
```{r}

```
#### Predictive analysis using KNN 
```{r}

```
#### Predictive analysis using Random Forest 
```{r}

```
#### Compare results of KNN and Random Forest and thus select a model 
```{r}

```
### Model Fine tuning

#### Fine tune the hyperparameters for the selected model ""
```{r}

```
#### Get results for the fine tuned model
```{r}

```
